DRAFT: Factors Influencing Abundance & Virus Isolation in Mosquitoes

### 2024-03-15 R/processing-code/MADAproject_pt2_ruiz.qmd
Part 3
Continued work on EDA. Most work for part 3 is contained in the R/eda-code folder. R code will eventually be pulled into the qmd via rds files. Continuing to update the manuscript and add better descriptions to code to facilitate better replication. Still work in progress. All read me files have been updated and will need further attention as the project moves forward.

## Bird data
### The complete bird dataset is over 600 MB. Because it is too large for GitHub, it is stored locally. It contain variables and records that are not needed for this analysis. Eight of 50 varibles will be kept and the records will be filtered to show only birds from the Genus Corvus observed in Massachusetts after 2014-01-01
```{r}
# Load complete bird data
bird <- read.table("/Users/andrewruiz/0011508-240216155721649.csv", sep = "\t", header = TRUE, fill = TRUE)

@@

#load the first few rows of data to verify the data loaded as expected
head(wx)

#the columns below are the only ones needed for this project. 
wx_selected <- wx %>%
  select(STATION, NAME, DATE, PRCP)

# Define the path for the RDS file
output_path_wx <- here("data", "processed-data", "rds", "wx_selected.rds")

# Save the wx_selected dataframe as an RDS file
saveRDS(wx_selected, output_path_wx)

head(wx_selected)
```

@@

#### The mosquito dataset spans 2007 to 2023. However, the virus and case data only covers the years 2014 to 2020. Because mosquito populations can be influenced by processes of the previous year. We will set the analytical horizon will span 2013 to 2021.*Collection data before 2013 and after 2021 will be removed* along with records with important missing variables identified below.

### Identify missing data
```{r}
sapply(mosquito_raw, function(x) sum(is.na(x)))
```

@@

# Add "week_num" column
# This code ensures that the ISO 8601 standard is applied to the week number calculation
# Week 1 of any year is the week that contains January 4th, or equivalently, it's the week that contains the first Thursday of January.
# Weeks start on Monday and end on Sunday.
# The last week of the year, week 52 or 53, is the one that contains December 28th.
# we will use week_num to join tables
# the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08
mosquito_modified <- mosquito_modified %>%
  mutate(
    # Directly format the collection date into "YYYY-WW"
    week_num = paste0(year(`collection date`), "-", 
                      sprintf("%02d", isoweek(`collection date`)))
  )

# before filtering the dates to have a cleaned dataset with all the records.
write_csv(mosquito_modified, here("data", "processed-data", "mosquito_clean_all.csv"))
saveRDS(mosquito_modified, here("data", "processed-data", "rds", "mosquito_clean_all.rds"))

# Now apply the date filter to match the analytical horizon of the project
mosquito_final <- mosquito_modified %>%
  filter(`collection date` >= as.Date("2013-01-01") & `collection date` <= as.Date("2021-12-31"))

# Check the structure of the final cleaned dataframe
str(mosquito_final)
```

## Reformat the collection data
### The Massachusetts Department of Public Health and US CDC require that all mosquito control districts submit their colelction data in a specific format. This format divides each trap event into separate rows based on mosquito species. However, this format is not useful for certain calculations where a zero count for a species in not explicitly recorded. Even if you sort by species and calculate average count per trap, the average could be inflated since the records will not include trap events with a zero count of that species.
### In order to correct this, I will pivot the data and create a column for every unique "species code" and sum the "pool size" for each species code. The resulting table will have one row for every trap event -a trap event is when "town", "date of collection", "trap type" are all the same.
### The resulting data frame will rectangle in shape and contain 1 row per trap event with all possible species as a separate column. 
```{r}
# Group and summarize the mosquito data to prepare for wide format transformation.
# This step aggregates the total 'pool size' for each unique combination of 
# 'collection date', 'town', 'trap type', 'species code', and 'mcd' (Mosquito Control District),
# while excluding 'submitted for testing' from the grouping to avoid potential inconsistencies.
mosquito_summarized <- mosquito_final %>%
  group_by(`collection date`, town, `trap type`, `species code`, mcd, `submitted for testing`) %>%
  summarise(pool_size_sum = sum(`pool size`, na.rm = TRUE), .groups = 'drop')
   # Ensure that the summarization does not drop any groups by setting .groups = 'drop'.

# Transform the summarized mosquito data to a wide format where each 'species code' becomes a separate column.
# This pivot operation facilitates easier analysis of mosquito counts by species across different events.
# 'values_fill' is used to fill in zeroes for any 'species code' not caught in a specific trapping event,
# ensuring that the dataset has a consistent shape and making it more analyzable.
wide_data <- mosquito_summarized %>%
  pivot_wider(
    names_from = `species code`,
    values_from = pool_size_sum,
    values_fill = list(pool_size_sum = 0),
    id_cols = c(`collection date`, town, `trap type`, mcd)
  )

@@

# Save the new file
# Specify the file path using here()
file_path_wide <- here("data", "processed-data", "mosquito_wide.csv")

@@

# Print the file path to confirm where the file has been saved
print(paste("File saved to:", file_path_wide))

# Define the path for the wide-format dataset RDS
file_path_wide_rds <- here("data", "processed-data", "rds", "mosquito_wide.rds")

@@

>>>>

## Bird data
### The complete bird dataset is over 600 MB. Because it is too large for GitHub, it is stored locally. It contain variables and records that are not needed for this analysis. Eight of 50 varibles will be kept and the records will be filtered to show only birds from the Genus Corvus observed in Massachusetts after 2014-01-01
```{r}
# Load complete bird data
# this uses an absolute path because the file is too large to store on github. 
# the filtered_bird_data.csv file will be the one used for any analysis
bird <- read.table("/Users/andrewruiz/0011508-240216155721649.csv", sep = "\t", header = TRUE, fill = TRUE)

@@

wx <- wx %>%
  mutate(
    DATE = mdy(DATE),  # Parse the DATE column from "month/day/year"
    week_num = paste0(year(DATE), "-", sprintf("%02d", isoweek(DATE)))  # Create week_num
  )
#load the first few rows of data to verify the data loaded as expected
head(wx)

#the columns below are the only ones needed for this project. 
wx_selected <- wx %>%
  select(STATION, NAME, DATE, PRCP, TMAX, week_num)

# Define the path for the RDS file
output_path_wx <- here("data", "processed-data", "rds", "wx_selected.rds")
output_csv_wx <- here("data", "processed-data", "wx_selected.csv")
# Save the wx_selected dataframe as an RDS file
saveRDS(wx_selected, output_path_wx)
write.csv(wx_selected, output_csv_wx, row.names = FALSE)
head(wx_selected)
```

@@

#### The mosquito dataset spans 2007 to 2023. However, the virus and case data only covers the years 2012 to 2020. We will create two datasets One that filters the dates to cover the virus and case data. And one with all records. 
### Identify missing data
```{r}
sapply(mosquito_raw, function(x) sum(is.na(x)))
```

@@

# Add "week_num" column
# This code ensures that the ISO 8601 standard is applied to the week number calculation
# Week 1 of any year is the week that contains January 4th, or equivalently, it's the week that contains the first Thursday of January.
# Weeks start on Monday and end on Sunday.
# The last week of the year, week 52 or 53, is the one that contains December 28th.
# we will use week_num to join tables
# the format will be YYYY-WW. Week 8 of 2019 will look like: 2019-08
mosquito_modified <- mosquito_modified %>%
  mutate(
    week_num = paste0(year(`collection date`), "-", sprintf("%02d", isoweek(`collection date`)))
  )
# Check the structure of the final cleaned dataframe
str(mosquito_modified)

# before filtering the dates to have a cleaned dataset with all the records.
write_csv(mosquito_modified, here("data", "processed-data", "mosquito_clean_all.csv"))
saveRDS(mosquito_modified, here("data", "processed-data", "rds", "mosquito_clean_all.rds"))
```

## Reformat the collection data
### The Massachusetts Department of Public Health and US CDC require that all mosquito control districts submit their colelction data in a specific format. This format divides each trap event into separate rows based on mosquito species. However, this format is not useful for certain calculations where a zero count for a species in not explicitly recorded. Even if you sort by species and calculate average count per trap, the average could be inflated since the records will not include trap events with a zero count of that species.
### In order to correct this, I will pivot the data and create a column for every unique "species code" and sum the "pool size" for each species code. The resulting table will have one row for every trap event -a trap event is when "town", "date of collection", "trap type" are all the same.
### The resulting data frame will rectangle in shape and contain 1 row per trap event with all possible species as a separate column. 
```{r}
# Group and summarize the mosquito data to prepare for wide format transformation.
# This step aggregates the total 'pool size' for each unique combination of 
# 'collection date', 'town', 'trap type', 'species code', and 'mcd' (Mosquito Control District),
# while excluding 'submitted for testing' from the grouping to avoid potential inconsistencies.
mosquito_summarized <- mosquito_modified %>%
  group_by(`collection date`, town, `trap type`, `species code`, mcd, `submitted for testing`) %>%
  summarise(pool_size_sum = sum(`pool size`, na.rm = TRUE), .groups = 'drop')
   # Ensure that the summarization does not drop any groups by setting .groups = 'drop'.
mosquito_summarized <- mosquito_modified %>%
  group_by(`collection date`, week_num, town, `trap type`, `species code`, mcd, `submitted for testing`) %>%
  summarise(pool_size_sum = sum(`pool size`, na.rm = TRUE), .groups = 'drop')

# Transform the summarized mosquito data to a wide format where each 'species code' becomes a separate column.
# This pivot operation facilitates easier analysis of mosquito counts by species across different events.
# 'values_fill' is used to fill in zeroes for any 'species code' not caught in a specific trapping event,
# ensuring that the dataset has a consistent shape and making it more analyzable.
wide_data <- mosquito_summarized %>%
  pivot_wider(
    names_from = `species code`,
    values_from = pool_size_sum,
    id_cols = c(`collection date`, week_num, town, `trap type`, mcd)
  ) %>%
  # Now use replace_na() to fill in NAs, specifying columns dynamically or by name after checking what columns are created
  replace_na(list(your_dynamic_or_specific_column_names_here = 0))

@@

# Save the new file
# Specify the file path using here()
file_path_wide <- here("data", "processed-data", "mosquito_wide_all.csv")

@@

# Define the path for the wide-format dataset RDS
file_path_wide_rds <- here("data", "processed-data", "rds", "mosquito_wide_all.rds")

@@

# Now we will filter the the mosquito data to match the time period of the virus data. 
# Since mosquito populations can be affected by the previous year's populations, we will filter for 2011-

# Filter wide_data for the specified date range
filtered_mosquito_data <- wide_data %>%
  filter(`collection date` >= as.Date("2011-01-01") & `collection date` <= as.Date("2022-12-31"))

# Define the path for the CSV file
file_path_filtered_csv <- here("data", "processed-data", "mosquito_wide_filtered.csv")

# Save the filtered wide-format data to the specified CSV path
write_csv(filtered_mosquito_data, file_path_filtered_csv)

# Define the path for the RDS file
file_path_filtered_rds <- here("data", "processed-data", "rds", "mosquito_wide_filtered.rds")

# Save the filtered wide-format data as an RDS file
saveRDS(filtered_mosquito_data, file_path_filtered_rds)

### 2024-03-15 products/manuscript/Manuscript.qmd
Part 3
Continued work on EDA. Most work for part 3 is contained in the R/eda-code folder. R code will eventually be pulled into the qmd via rds files. Continuing to update the manuscript and add better descriptions to code to facilitate better replication. Still work in progress. All read me files have been updated and will need further attention as the project moves forward.

Eastern equine encephalitis (EEE) is caused by an Alphavirus transmitted
to humans by the bite of an infected mosquito. Human infections are
rare, but serious. About 30-50% of symptomatic cases lead to death and
those who survive are left with life-long disability. The transmission
cycle for EEE includes two distinct cycles: an amplifying enzootic cycle
in which the virus is transmitted between the Culiseta melanura mosquito
and birds and an epizootic cycle in which bridge vector mosquitoes
transmit the EEE virus (EEEv) from birds to humans and other mammals.
Because C. melanura feeds almost exclusively on birds, it is not
considered a direct human threat. However, EEEv isolations in C.
melanura are an early warning sign that EEEv is circulating in the
ecosystem. Isolations in bridge vector mosquito species indicate
heightened transmission risk to humans.

@@

This project will lay the foundation for dissertation research that will
begin in June, 2024. The research will focus on three

# Introduction

## Background

Humans and other mammals are dead-end hosts for EEE. About 94% of human
EEE infections are asymptomatic. While symptomatic cases are rare, the
consequences can be severe. Mortality rates associated with symptomatic
infections range from 30 to 50%. Those who survive often experience
lifelong disability that includes debilitating neurological damage. On
average, there are 11 cases reported in the US annually. Most occur in
Massachusetts and Florida. Until recently, nearly all human EEE cases in
Massachusetts occurred in two counties: Bristol and Plymouth.

@@

# Methods

## Study area

Bristol County, Massachusetts covers an area of 691 square miles with a
population density of 962 people per square mile. It borders Rhode
Island to the west, Plymouth County, MA to the east, and Norfolk County,
MA to the north.The county is situated on Buzzard's Bay to the south.

Atlantic White Cedar (AWC) swamps span over 4,300 acres within
Massachusetts, primarily concentrated in Plymouth and Bristol counties.
These swamps serve as critical breeding habitats for *Culiseta melanura*
(MEL), which is recognized as the primary enzootic vector for Eastern
Equine Encephalitis (EEE). AWC trees have large roots that form "crypts"
that hold water throughout the year. These crypts serve as a main
habitat for MEL larval growth.

@@

## Data collection and processing

The mosquito surveillance dataset encompasses counts by species from all mosquito traps deployed in Bristol County, MA, from 2007 to 2023. The BCMCP, perform mosquito surveillance from June through October annually. During this interval, CDC light traps baited with carbon dioxide are placed at various sentinel sites each week. Some of the sentinel site have been used for over 50 years. The traps are installed in the afternoon and retrieved the following morning. Designed to attract mosquitoes in search of blood meals, light traps serve as a reliable measure of the actively feeding mosquito population, thereby establishing them as the standard method for EEE surveillance.

The  In addition to mosquito
counts by species, the dataset includes the town or city in which they
were collected, the collection data, and whether the mosquitoes were
submitted for virus testing.

While the surveillance data provide insight into mosquito population
estimates and changes in species composition, it is missing one key
feature. The virus isolation results are not included in this dataset.
However, virus isolations in mosquitoes are published on the MDPH
website by town/city and collection date. With this information, one can
identify the trap event contained in the dataset which allow for
calculations of key measure, such as minimum infection rate (MIR).

This dataset will be integrated with other datasets using spatial or
temporal attributes. These datasets will include biological and
ecological data, environmental and spatial data, and demographic data,
which are crucial in understanding the transmission dynamics of EEE.
Potential datasets with their sources are outlined in the table below.
Other datasets may be included based on availability and need.

## Questions/Hypotheses to be addressed

\*This project will lay the foundation for my dissertation research.
**The goal for this phase of the project is to answer which
environmental, ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts.** From prior
research we know that the mosquito life cyle is influenced by a number
of factors. Temperature plays a major role in the mosquito life cyle
virus transmission dynamics. As temperature increases, the rate of
larval development also increases and the time between when a mosquito
ingests virus in a bloodmeal and when it can transmit it decreases. The
result is that the mosquito population can grow at faster rate and they
will have more time to transmit the virus. Precipitation increases the
amount of standing water and increases the chances of larval survival
during the mosquito season and also during the winter months for those
species that overwinter in the larval stage. For EEE, the main enzootic
vector, MEL, overwinters as larvae.

@@

{{< pagebreak >}}

# Methods

*Specific methods will be determined when all data sources are
confirmed.*

## The process will involve creating a relational database in which all datasets are joined using spatial or temporal attributes. This will lead to a thorough examination of the datasets to understand the distributions, identify missing data and outliers, and uncover potential correlations between variables. The EDA will involve using visualizations, geographic information systems, and statistical tests to understand how biological, ecological, environmental, and demographic forces influence the EEE transmission cycles. The EDA will inform the preprocessing steps, feature engineering, and selection of appropriate AI models.

### Schematic of workflow

*This section will be updated before analysis begins*

## Data aquisition

With the exception of the mosquito collection data, all other datasets
are publicly available for use without restrictions.

## Data import and cleaning

Datasets will be loaded into R in a Quarto document to document and
describe cleaning methods. There are two datasets that will require more
extensive cleaning: the mosquito trap data and the crowd-sourced bird
data. The process is outlined in detail in the MADAproject_pt2.qmd file.
Less significant cleaning and data management steps not included in this
manuscript are also outlined in that file.

The mosquito dataset spans 2007 to 2023. However, the virus and case
data only covers the years 2014 to 2020. Because mosquito populations
can be influences by processes of the previous year. We will set the
analytical horizon will span 2013 to 2021.

Additionally, the Massachusetts Department of Public Health requires
that all mosquito control district submit their colelction data in a
specific format. This format divides each trap event into separate rows
based on mosquito species. However, this format is not useful for
certain calculations where a zero count for a species is not explicitly
recorded. For example, even when sorted by species to calculate the
average count per trap, the average could be inflated since the records
will not include trap events with a zero count of that species. In order
to correct this, the data was pivoted to create a column for every
unique "species code" then the "pool size" for each species code is
summed for each trap event. A trap event is when "town", "date of
collection", "trap type" are all the same.. For species codes not
included in the orginal table for a trap event, the field will be
assigned a zero value. The resulting table will have one row for every
trap event.

The bird observation dataset was obtained from Cornell University's
eBird program. The original dataset included all reported bird
observations for Southeast Massachusetts and was over 600 MB of data.
While all birds are potential amplifying hosts for EEE and West Nile
Virus (WNV), birds from the Genus *Corvus* are of particular concern.
The bird dataset processed to include records of *Corvus spp* sightings
after 2014. The final dataset included 8 of the 50 variables, including
location data.

## Statistical analysis

This exercise will serve as the beginning of the exploratory data
analysis phase. One of the main objectives for of this phase is describe
the data, understand its structure, clean data, and identify
associations within the datasets.

# Results

## Exploratory/Descriptive analysis

The first phase of the exploration is complete. The next phase will
include statistical tests to determine if there are possible
associations between the environmental variables and the mosquito
abundance as measured by trap counts or EEE virus activity.

@@

The geographic focus for this study is Bristol County. One factor to
consider is if early virus isolations are associated with cases in
humans and other mammals. The table below shows the date of the first
EEE virus isolation in mosquitoes and the number of human of other
animal cases by year.

@@

## Basic statistical analysis

*These will be generated in later phases of the exercise*

@@

*this section will be updated after analysis is complete*

While the reliability of the mosquito trap data is high, the trap data
cannot be directly associated with the virus isolation data. The
information this scale is not released by the State. Spatial scale is
also an issue with the human and veterinary cases data. Human case data
is released only on a county scale to ensure patient privacy. Some
veterinary case data is available on the town-level, but most remains
only on the county-level.

These limitations are barriers to accurate spatial analysis.

@@

References will be added in part 3 of the project

>>>>

*This project will inform the development of my dissertation prospectus
by identifying significant relationships between variables and guide the
methods development.* 
Eastern equine encephalitis (EEE) is caused by an
Alphavirus transmitted to humans by the bite of an infected mosquito.
Human infections are rare, but serious. About 30-50% of symptomatic
cases lead to death and those who survive are left with life-long
disability. The transmission cycle for EEE includes two distinct cycles:
an amplifying enzootic cycle in which the virus is transmitted between
the Culiseta melanura mosquito and birds and an epizootic cycle in which
bridge vector mosquitoes transmit the EEE virus (EEEv) from birds to
humans and other mammals. Because C. melanura feeds almost exclusively
on birds, it is not considered a direct human threat. However, EEEv
isolations in C. melanura are an early warning sign that EEEv is
circulating in the ecosystem. Isolations in bridge vector mosquito
species indicate heightened transmission risk to humans.

@@

**This project will lay the foundation for dissertation research that will
begin in June, 2024.**

# Introduction
## Background
*Citations will be added when the document is finalized* 
Humans and other mammals are dead-end hosts for EEE. About 94% of human EEE
infections are asymptomatic. While symptomatic cases are rare, the
consequences can be severe. Mortality rates associated with symptomatic
infections range from 30 to 50%. Those who survive often experience
lifelong disability that includes debilitating neurological damage. On
average, there are 11 cases reported in the US annually. Most occur in
Massachusetts and Florida. Until recently, nearly all human EEE cases in
Massachusetts occurred in two counties: Bristol and Plymouth.

@@

## Study area
Bristol County, Massachusetts covers an area of 691 square miles with a
population density of 962 people per square mile. It borders Rhode
Island to the west, Plymouth County, MA to the east, and Norfolk County,
MA to the north. Naushon Island and Martha's Vineyard are across
Buzzard's Bay to the Southeast.

Atlantic White Cedar (AWC) swamps span over 4,300 acres within
Massachusetts, primarily concentrated in Plymouth and Bristol counties.
These swamps serve as critical breeding habitats for *Culiseta melanura*
(MEL), which is recognized as the primary enzootic vector for Eastern
Equine Encephalitis (EEE). AWC trees have large root systems that form
"crypts" and hold water throughout the year. These crypts serve as a
main habitat for MEL larval growth and provide shelter during winter
months.

@@

## Questions/Hypotheses to be addressed
*This project will lay the foundation for my dissertation research.*
**The goal for this phase of the project is to answer which
environmental, ecological, and temporal factors influence the
transmission cycles for EEE in southeastern Massachusetts.** From prior
research we know that the mosquito life cyle is influenced by a number
of factors. Temperature plays a major role in the mosquito life cyle
virus transmission dynamics. As temperature increases, the rate of
larval development also increases and the time between when a mosquito
ingests virus in a bloodmeal and when it can transmit it decreases. The
result is that the mosquito population can grow at faster rate and they
will have more time to transmit the virus. Precipitation increases the
amount of standing water and increases the chances of larval survival
during the mosquito season and also during the winter months for those
species that overwinter in the larval stage. For EEE, the main enzootic
vector, MEL, overwinters as larvae.

@@

## Data collection and processing
The mosquito surveillance dataset encompasses counts by species from all
mosquito traps deployed in Bristol County, MA, from 2007 to 2023. The raw dataset includes over 45,000 records and includes temporal and spatial variable. The BCMCP, perform mosquito surveillance from June through October annually.

During this interval, CDC light traps baited with carbon dioxide are
placed at various sentinel sites each week. Designed to attract
mosquitoes in search of blood meals, light traps serve as a reliable
measure of the actively feeding mosquito population, thereby
establishing them as the standard method for EEE surveillance. Some of
the sentinel site have been used for over 50 years.

The traps are installed in the afternoon and retrieved the following
morning. After the traps are collected, the next are stored in a freezer
to preserve and kill the mosquitoes. The mosquitoes are sorted by
species using a dissecting scope. Counts for each species are recorded
in the database and a portion of the vector species are driven to the
Massachusetts State Laboratory in Boston. The lab tests each mosquito
pool (or batch) for WNV and EEEv using PCR. Results are shared with
BCMCP on the same day of testing. Virus isolation results are recorded
in the collection records.

In addition to mosquito counts by species, the dataset includes the town
or city in which they were collected, the collection date, and whether
the mosquitoes were submitted for virus testing.

One indicator of the level of virus activity present in the mosquito
population can be calculated from the information contained in dataset.
The minimum infection rate (MIR) can be calculated at various scales
from trap site to all sites over an entire season. MIR is calculated as:
$$
MIR = \left( \frac{\text{Number of positive mosquito pools}}{\text{Total number of mosquitoes tested}} \right) \times 1000
$$ 
MIR provides an estimate of the prevalence of a specific pathogen
within a mosquito population. While it does not directly measure the
risk to humans, a higher MIR in mosquitoes that commonly bite humans can
indicate a higher risk of transmission. MIR allows for the comparison of
infection rates across different geographical areas or time periods.
This is crucial for identifying areas or times of higher risk and for
evaluating the effectiveness of control measures.

![Minimum infection rate by
week](images/MIR_by_week_heatmap.png){fig-align="center"}

Weather data played a significant role in this phase of the analysis. Daily weather summaries were downloaded from NOAA's Climate Data Portal. The dataset covers the years 2007-2023 at the Taunton Airport in Bristol County.

These datasets will be integrated with other datasets using spatial or
temporal attributes. These datasets may include biological and
ecological data, environmental and spatial data, and demographic data,
which are crucial in understanding the transmission dynamics of EEE.
Potential datasets with their sources are outlined in the table below.
Other datasets may be included based on availability and need.

## Data aquisition
With the exception of the mosquito collection data, all other datasets
are publicly available for use without restrictions.

## Data import and cleaning
Datasets will be loaded into R in a Quarto document to document and
describe cleaning methods. There are two datasets that will require more
extensive cleaning: the mosquito trap data and the crowd-sourced bird
data. The process is outlined in detail in the MADAproject_pt2.qmd file.
Less significant cleaning and data management steps not included in this
manuscript are also outlined in that file.

*I am deciding how to handle the time period. I do have good data for some variables that span the entire period.*
The mosquito dataset spans 2007 to 2023. However, the virus and case
data only covers the years 2014 to 2020. Because mosquito populations
can be influences by processes of the previous year. We will set the
analytical horizon will span 2013 to 2021.

Additionally, the Massachusetts Department of Public Health requires
that all mosquito control district submit their collction data in a
specific format. This format divides each trap event into separate rows
based on mosquito species. However, this format is not useful for
certain calculations where a zero count for a species is not explicitly
recorded. For example, even when sorted by species to calculate the
average count per trap, the average could be inflated since the records
will not include trap events with a zero count of that species. In order
to correct this, the data was pivoted to create a column for every
unique "species code" then the "pool size" for each species code is
summed for each trap event. A trap event is when "town", "date of
collection", "trap type" are all the same.. For species codes not
included in the orginal table for a trap event, the field will be
assigned a zero value. The resulting table will have one row for every
trap event.

The bird observation dataset was obtained from Cornell University's
crowd sourced eBird program. The original dataset included all reported bird
observations for Southeast Massachusetts and was over 600 MB of data.
While all birds are potential amplifying hosts for EEE and West Nile
Virus (WNV), birds from the Genus *Corvus* are of particular concern.
The bird dataset processed to include records of *Corvus spp* sightings
after 2014. The final dataset included 8 of the 50 variables, including
location data. *this dataset is of poor quality. I am looking for a
better dataset now. Looking into the annual breeding bird survey*

# Methods
*This section is under development and will be completed after analysis
is complete. The following is a general outline of the methods.* All
datasets with a temporal attribute will be modifided to include a new
for the week number. Week number calculations will be standardized
according to ISO 8601 standards and will be stored an a YYYY-WW format.
For example, week 8 in 2019 will be displayed as 2019-08. Week numbers
will provide a common field by which all datasets with a temporal
attribute can be joined. It will also allow for grouping records and
making calculations by week number. This format helps to standardize
temporal scales and make comparisons between years easier.

Symptomatic EEE, and to a lesser extent WNV, human cases are rare. On
average, the United States records only about 9 cases of EEE annually.
Notably, Massachusetts reports more cases than any other state except
Florida, highlighting regional disparities in disease occurrence.
However, the incidence rate for neuroinvasive EEE cases in Bristol
County, Massachusetts, is estimated to be only around 0.06 per 100,000
people. The infrequency of these cases poses significant challenges for
traditional statistical models, making it difficult to accurately
estimate human transmission risk. This difficulty is largely due to the
sparse nature of the data and the variability in transmission patterns,
which complicates efforts to predict outbreaks and necessitates
alternative modeling approaches.

### Schematic of workflow
*Schematic will be refined in later steps. The doc will also be
reformatted to move the schematic to an appendix* *The image should
appear upon rendering*

```{mermaid}
flowchart TB
    id1["Questions"] --> id2["Data Collection"]
    id2 --> id3["Data Loading and Cleaning"]
    id3 --> id4["Data Cleaning"]
    id4 --> id5["Data Transformation"]
    id5 --> id6["Data Exploration & Visualization"]
    id6 --> id7["Data Modeling"]
    id7 --> id8["Evaluation"]

    id8 -.-> id1
    id8 --> id9[Interpretation]
    id9 --> id10[Communication]

    classDef mainProcess fill:#fff,stroke:#333,stroke-width:2px;

    class id1,id2,id3,id4,id5,id6,id7,id8,id9,id10 mainProcess
```

## Exploratory/Descriptive analysis
The first phase of the exploration is complete. The next phase will
include statistical tests to determine if there are possible
associations between the environmental variables and the mosquito
abundance as measured by trap counts or EEE virus activity.

@@

One factor to consider is if early virus isolations are associated with cases in
humans and other mammals. The table below shows the date of the first
EEE virus isolation in mosquitoes and the number of human of other
animal cases by year.

@@

## Basic statistical analysis
*More will be generated in later phases of the exercise* The initial
tests have been basic regression models.

The first models explored the relationship between virus isolation in
mosquitoes and weather conditions (average temperature by week number
and total precipitation by week). The outcome was whether there was at
least 1 positive pool in a week. The predictor in the first model was
the average maximum daily temperature the week before the pool week. The
logistic model suggest that there was a relationship between the two. (The R code for the model can be found in the EDA-Code file: logistic_.pos_temp_precip.R) The average temperature in the week before is positively associated with
the likelihood of a positive result, with an odds ratio of approximately
1.069 for each unit increase in temperature (Estimate = 0.06695, SE =
0.02033, p \< 0.001), indicating that higher temperatures significantly
increase the probability of a positive outcome. The model's improvement
is highlighted by a decrease in residual deviance to 405.79 from the
null deviance of 417.24, upon incorporating Lagged_Avg_TMAX as a
predictor, demonstrating a better fit to the observed data. However,
with a residual deviance of 405.79 on 299 degrees of freedom, it is
evident that not all variability in the outcome is explained, suggesting
the presence of other influential factors. The model's AIC of 409.79
indicates a balance between model complexity and predictive power,
supporting its relative utility in understanding the data. However, this relationship should be examined more closely as temperature is likely auto correlated with progressing weeks during summer months. While
precipitation has been demonstrated to influence the population dynamics
it did not show to have any effect on a positive result.

Another model assessed whether the temperature precipitation in April, May, and June of each year effected the date of the first positive pool test. The date of the first positive pool was identified and assigned a day number from January 1st of that year. The predictors in this linear model were the average  maximum temperature for April, May, and June and the sum of precipitation for the same time period. None of these predictors were significant.

There was one other significant result. When MIR is set as the dependent, continuous variable in a linear regression and the average maximum temperature is used as the predictor, the model suggests there is a significant inverse relationship. As average temperature increases, the MIR decreases (model_mir_temp.R). This relationship is counterintutive, especially given the result of the logitic regression and requires more examination.

More models assessing the effects of weather and other factors on positive pools by mosquito species and MIR values will be developed in Part 4 of the assignment. Future models will examine the the effects of weather conditions on mosquito population dynamics.

## Statistical analysis
This exercise will serve as the beginning of the exploratory data
analysis phase. One of the main objectives for of this phase is describe
the data, understand its structure, clean data, and identify
associations within the datasets.

# Results
*updated in later phases of the project*

@@

*this section will be updated after analysis is complete* The accuracy
and reliability of the mosquito collection data is excellent. The same
procedures and sentinel sites have were in place for the entire
analytical horizon. Further, there were only two people responsible for
sorting, identifying and counting species over the entire period.

MIR is not a perfect measure of virus activity in mosquitoes. It tends
to underestimate the level of virus circulating in the environment.
However, other measures require equal pool numbers for testing. There is
a correction factor for variable pool numbers and I will explore this
option in later phases of this project.

The spatial scale is an issue with the human and veterinary cases data.
Human case data is released only on a county scale to ensure patient
privacy. Some veterinary case data is available on the town-level, but
most remains only on the county-level. These limitations are barriers to
accurate spatial analysis. The project will rely on virus isolation in
mosquitoes to model risk. The available bird data relies heavily on
crowd-sourced reports. There is one standardized survey (Breeding Bird
Survey) that seems to be more accurate and reliable. However, this data
is not available for every year.

@@

References will be added in part 4 of the project
